---
title: "HeartDiseaseNEW.RMD"
output: html_document
date: "2025-03-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preliminary Imports + Libraries

```{r}
library(glmnet)  #For regression
library(glm2) 	#Logistic Regression
library(ggplot2)  # For better plotting services
library(GGally)   # For a better scatterplot matrix ggpairs()
library(leaps)    # For stepwise and all-subsets
library(lmSubsets) 
library(MASS)     # Many useful statistics functions, matrix operartions, Linear Discriminant Analysis (LDA)
library(corrplot) # For a correlation plot
library(car)      # For vif
library(dplyr)
library(readxl)
library(factoextra)
library(caret) 	#Random Forests
library(ca)	#Correspondence Analysis
library(ROCR)	#ROC Curve + AUC
library(StatMatch)	#Gower Distance
library(dbscan)	#Density Based Clustering
library(cluster)		#Divergent (Backwards Hierarchical) Clustering
library(stats)  #In base R, PCA
library(dplyr)
library(polycor)   # For polychoric & polyserial correlations
library(cluster)   # For Gower's Distance (clustering)
library(psych)     # For Factor Analysis
library(FactoMineR) # For MCA (Multiple Correspondence Analysis)
library(factoextra) # For PCA/MCA visualization
library(ca)        # For Correspondence Analysis



heartDis <- read.csv("~/Desktop/DSC 324/Final Project/Heart_disease_cleveland_new.csv")
View(heartDis)
str(heartDis)

setwd("~/Desktop/DSC 324/Final Project")
```

# Pre-Processing

### Missing Values (None)

```{r}
# Check for missing values in each column
missing_values <- colSums(is.na(heartDis))

# Convert to a data frame for better readability
missing_values_df <- data.frame(Variable = names(missing_values), Missing_Count = missing_values)

# Print summary of missing values
print(missing_values_df)

```

### Plot variable distributions –\> outliers?

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Identify Numeric Variables
numeric_vars <- heartDis %>%
  select(where(is.numeric))

# Identify Categorical Variables
categorical_vars <- heartDis %>%
  select(where(is.factor))

# Function to plot numeric variable distributions
plot_numeric <- function(var) {
  print(
    ggplot(heartDis, aes_string(x = var)) +
      geom_histogram(bins = 30, fill = "blue", alpha = 0.6, color = "black") +
      ggtitle(paste("Distribution of", var)) +
      theme_minimal()
  )
}

# Function to plot categorical variable distributions
plot_categorical <- function(var) {
  print(
    ggplot(heartDis, aes_string(x = var)) +
      geom_bar(fill = "blue", alpha = 0.6, color = "black") +
      ggtitle(paste("Distribution of", var)) +
      theme_minimal() 
  )
}

# Plot numeric variables
lapply(names(numeric_vars), plot_numeric)

# Plot categorical variables
lapply(names(categorical_vars), plot_categorical)

###################################################################################################################
#Grid plotting, needs to be viewed by clicking plot symbol box in top right of RMD output window
###################################################################################################################

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(gridExtra)

# Identify Numeric Variables
numeric_vars <- heartDis %>%
  select(where(is.numeric))

# Identify Categorical Variables
categorical_vars <- heartDis %>%
  select(where(is.factor))

# Function to plot numeric variable distributions
plot_numeric <- function(var) {
  ggplot(heartDis, aes_string(x = var)) +
    geom_histogram(bins = 30, fill = "blue", alpha = 0.6, color = "black") +
    ggtitle(paste("Distribution of", var)) +
    theme_minimal()
}

# Function to plot categorical variable distributions
plot_categorical <- function(var) {
  ggplot(heartDis, aes_string(x = var)) +
    geom_bar(fill = "blue", alpha = 0.6, color = "black") +
    ggtitle(paste("Distribution of", var)) +
    theme_minimal()
}

# Generate plots for numeric variables
numeric_plots <- lapply(names(numeric_vars), plot_numeric)

# Generate plots for categorical variables
categorical_plots <- lapply(names(categorical_vars), plot_categorical)

# Combine all plots
all_plots <- c(numeric_plots, categorical_plots)

# Display all plots in a grid layout
grid.arrange(grobs = all_plots, ncol = 3)  # Adjust 'ncol' to change layout

```

### Detecting and Removing Outliers:

-   oldpeak \>4 , chol \> 450, trestbps \>180

-   8 observations removed

```{r}
#############################################################################################
# Detect and Print Outliers Before Removal
#############################################################################################

# Load necessary library
library(dplyr)

# Detect outliers based on given conditions
outliers_detected <- heartDis %>%
  mutate(ID = row_number()) %>%  # Add row ID for reference
  filter(oldpeak > 4 | trestbps > 180 | chol > 450)

# Print detected outliers with row ID
cat("\n📌 **Detected Outliers Before Removal** 📌\n")
print(outliers_detected)

# Count the number of detected outliers
cat("\n✅ Number of detected outliers:", nrow(outliers_detected), "\n")

#############################################################################################
# Remove Specified Outliers
#############################################################################################

# Remove outliers based on criteria
heartDis_cleaned <- heartDis %>%
  filter(oldpeak <= 4, 
         trestbps <= 180, 
         chol <= 450)

# Print summary of the dataset after removal
cat("\n📌 **Summary After Outlier Removal** 📌\n")
print(summary(heartDis_cleaned))

# Count the number of rows removed
rows_removed <- nrow(heartDis) - nrow(heartDis_cleaned)
cat("\n✅ Number of rows removed:", rows_removed, "\n")

str(heartDis)
str(heartDis_cleaned)



```

### Define Variable Types

```{r}
# Convert Binary variables to factors
binary_vars <- c("sex", "fbs", "exang")   # Binary (0/1)
heartDis_cleaned[binary_vars] <- lapply(heartDis_cleaned[binary_vars], as.factor)

# Convert Nominal variables to factors
ordinal_vars <- c("cp", "restecg", "slope", "ca", "thal")   # Ordinal (ordered categories)
heartDis_cleaned[ordinal_vars] <- lapply(heartDis_cleaned[ordinal_vars], as.factor)

# Ensure Numeric variables are numeric
numeric_vars <- c("age", "trestbps", "chol", "thalach", "oldpeak")  # Continuous numeric
heartDis_cleaned[numeric_vars] <- lapply(heartDis_cleaned[numeric_vars], as.numeric)

#Check structure to ensure proper transformations
str(heartDis_cleaned)

# Select predictor variables, excluding 'target'
pred_vars <- heartDis_cleaned %>%
  select(-target)

```

### Compute Mixed Correlation Matrix

-   This will generate a correlation matrix that can be used for **Factor Analysis or PCA**.

-   **(Polyserial + Polychoric + Pearson) =** cor_matrix

```{r}
# Compute hybrid correlation matrix for numeric, ordinal, and binary variables
cor_matrix <- hetcor(heartDis_cleaned[, c(numeric_vars, ordinal_vars, binary_vars)])$correlations

# Print mixed correlation matrix
cat("\n **Mixed Correlation Matrix** \n")
print(cor_matrix)

# Plot the correlation matrix using corrplot
corrplot(cor_matrix, method = "color", title = "Mixed Correlation", mar = c(0,0,1,0))

```

### Parallel Analysis (Number of Factors)

Specify the Number of Components to get an initial idea of the interpretability of the components

-   4-5 components is necessary

```{r}
#Parallel Analysis Scree Plots
  #Parallel analysis does Factore analysis and PCA side by side
Plel <- fa.parallel(cor_matrix)
Plel

```

### Exploratory Factor Analysis (EFA) with Varimax

```{r}
# Run the Exploratory Factor Analysis
# Here we use Maximum Likelihood extraction (fm = "ml") and Varimax rotation.
# Adjust nfactors to the number of factors you want to extract (e.g., 3).
efa_result <- fa(r = cor_matrix, nfactors = 5, fm = "ml", rotate = "Varimax")

# Print the EFA results summary
print(efa_result)

# Optionally, display factor loadings with a cutoff (e.g., 0.3) for easier interpretation
cat("\n EFA with all Variables \n")
print(efa_result$loadings, oreder=T, cutoff = 0.5)
```

### KMO and Bartlett - Suitability of data for FA

-   KMO and Bartlett

-   Compare with CFA factanal() after

<!-- -->

-   Bartlett p-val \<2e-16 –\> We are suitable for FA

-   Overall KMO: 0.702–\> We are suitable for FA

    -   Could potentially drop 5/13 predictors:
        -   **sex(0.424), chol(0.478), restecg(0.5), trestbps(0.517), and fbs(0.55)**

```{r}
# Perform Bartlett’s Test of Sphericity
bartlett_test <- cortest.bartlett(cor_matrix, n = nrow(heartDis_cleaned))

# Perform the KMO test
kmo_result <- KMO(cor_matrix)

# Print results with labels
cat("\n **Bartlett’s Test of Sphericity** \n")
cat("Chi-Square Value:", round(bartlett_test$chisq, 3), "\n")
cat("Degrees of Freedom:", bartlett_test$df, "\n")
cat("p-value:", format.pval(bartlett_test$p.value, digits = 3), "\n")

cat("\n **Kaiser-Meyer-Olkin (KMO) Measure of Sampling Adequacy** \n")
cat("Overall KMO:", round(kmo_result$MSA, 3), "\n")

cat("\n **KMO Individual Measures** \n")
print(round(kmo_result$MSAi, 3))

#############################################################################################
#Remove vars
#############################################################################################
# Subset pred_vars to remove 'sex','fbs','chol' and 'trestbps'
  #All weak predictors and have KMO < 0.65
# Define variables to remove
vars_to_remove <- c("sex", "chol", "trestbps", "restecg", "fbs")

# Combine your variable groups and remove the unwanted ones
selected_vars <- setdiff(c(numeric_vars, ordinal_vars, binary_vars), vars_to_remove)

cat("\n############## PREDS REMOVED #################\n")

#############################################################################################
#Re-Run with vars removed
#############################################################################################
# Compute the correlation matrix
# Compute the hybrid correlation matrix for the selected variables
cor_matrix5 <- hetcor(heartDis_cleaned[, selected_vars])$correlations

# Perform Bartlett’s Test of Sphericity
bartlett_test <- cortest.bartlett(cor_matrix5, n = nrow(heartDis_cleaned))

# Perform the KMO test
kmo_result <- KMO(cor_matrix5)

# Print results with labels
cat("\n **Bartlett’s Test of Sphericity** \n")
cat("Chi-Square Value:", round(bartlett_test$chisq, 3), "\n")
cat("Degrees of Freedom:", bartlett_test$df, "\n")
cat("p-value:", format.pval(bartlett_test$p.value, digits = 3), "\n")

cat("\n **Kaiser-Meyer-Olkin (KMO) Measure of Sampling Adequacy** \n")
cat("Overall KMO:", round(kmo_result$MSA, 3), "\n")

cat("\n **KMO Individual Measures** \n")
print(round(kmo_result$MSAi, 3))

```

### 

# Analysis 1: EFA fm = "minres" (Error)

-   No Varimax, fm = "ml"

<!-- -->

-   5 factors, 0.5 cutoff, fm = "ml"

```{r}
#X factors
  #Setting cor=T is essentially the same as using scale() befrehand because it
  #uses the correlation matrix rather than the covariance matrix
fitX <- fa(r= cor_matrix, nfactors= 5, fm = "minres", rotate ="varimax")

cat("\n EFA by Minimum Residuals with all varoiables \n")
print(fitX$loadings, order=T, cutoff= 0.5, sort= T)

cat("/n #############################################################/n")

```

# Analysis 2: CFA fm = "ml"

-   No Varimax, fm = "ml"

<!-- -->

-   5 factors, 0.5 cutoff, fm = "ml"

```{r}
#X factors
  #Setting cor=T is essentially the same as using scale() befrehand because it
  #uses the correlation matrix rather than the covariance matrix
fitX <- fa(r= cor_matrix, nfactors= 5, fm = "ml", rotate ="none")

cat("\n Analysis 2: CFA with all Variables, no rotation \n")
print(fitX$loadings, order=T, cutoff= 0.5, sort= T)

cat("/n #############################################################/n")

```

# Analysis 3: CFA fm = "ml", Varimax

-   No Varimax, fm = "ml"

<!-- -->

-   5 factors, 0.5 cutoff, fm = "ml"

```{r}
#X factors
  #Setting cor=T is essentially the same as using scale() befrehand because it
  #uses the correlation matrix rather than the covariance matrix
fitX <- fa(r= cor_matrix, nfactors= 5, fm = "ml", rotate ="varimax")

cat("\n Analysis 3: Varimax CFA with all Variables \n")
print(fitX$loadings, order=T, cutoff= 0.5, sort= T)

cat("/n #############################################################/n")

```

# 

# Analysis 4: CFA, fm= "ml", Varimax, vars KMO \< 0.6 dropped

Re-Run PFA with **sex(0.424), chol(0.478), restecg(0.5), trestbps(0.517), and fbs(0.55)**

-   4 factors are needed

    -   PROMAX rotation yields best results

5 factors:

-   Variance increased from 0.612 to 0.766

-   thalach is still the only negative correlation

-   ca in two factors

```{r}
###############################################################
#Use the Correlation Matrix without those 5 + re-remove target
###############################################################
# Subset pred_vars to remove 'sex','fbs','chol' and 'trestbps'
  #All weak predictors and have KMO < 0.65
# Define variables to remove
vars_to_remove <- c("sex", "chol", "trestbps", "restecg", "fbs", "target")

# Combine your variable groups and remove the unwanted ones
  #Remove 5, have 8 left
selected_vars <- setdiff(c(numeric_vars, ordinal_vars, binary_vars), vars_to_remove)

# Compute the hybrid correlation matrix for the selected variables
cor_matrix5 <- hetcor(heartDis_cleaned[, selected_vars])$correlations

# Plot the correlation matrix using corrplot
corrplot(cor_matrix5, method = "color", title = "Mixed Correlation", mar = c(0,0,1,0))
###############################################################
#Re-Run PFA to examine differences
  #4-5 factors is ideal
###############################################################

#X factors
fitX <- fa(r=cor_matrix5, 
           nfactors=5, 
           fm = "ml", 
           rotate ="varimax",
           n.obs = nrow(heartDis_cleaned), 
           scores = "regression")

cat("\n**Analysis 4: Varimax CFA with 5 preds removed ** \n")
print(fitX$loadings, order=T, cutoff= 0.5, sort= T)


```

### psych CFA -–\> levaan CFA to Extract Factor Scores

-   To obtain factor scores, you need to run the factor analysis on the raw data directly, can't obtain factor scores from categorical variable fa() tho

-   Can't do factor scores for single variable factors

```{r}
# install.packages("lavaan")  # if not already installed
library(lavaan)

# Suppose these variables are ordinal (or binary, treated as ordinal)
ordered_vars <- c("slope", "cp", "ca", "thal", "exang")

library(lavaan)

# Define the CFA model using only the two-indicator factors
model_revised <- '
  ML5 =~ oldpeak + slope
  ML3 =~ cp + exang
  ML2 =~ age + ca
  
  # Allow factors to correlate freely
  ML5 ~~ ML3 + ML2
  ML3 ~~ ML2
'

# List of ordinal/binary variables (if applicable)
ordered_vars <- c("cp", "slope", "ca", "thal", "exang")  # Adjust based on your dataset

# Fit the CFA model using WLSMV estimator (for ordinal data)
fit_cfa_revised <- cfa(
  model_revised,
  data      = heartDis_cleaned,
  ordered   = ordered_vars,  # Specify which variables are ordinal
  estimator = "WLSMV"
)

# Check model summary
summary(fit_cfa_revised, fit.measures = TRUE, standardized = TRUE)

#############################################################################################
# Extract factor scores for ML5, ML3, ML2
factor_scores <- lavPredict(fit_cfa_revised)

# Convert to a dataframe and attach the observed variables
df_model <- cbind(factor_scores, 
                  thalach = heartDis_cleaned$thalach, 
                  thal = heartDis_cleaned$thal)

# Check structure
head(df_model)

```

### Run LASSO regression

```{r}
library(glmnet)

# Convert predictors to a matrix for glmnet
X <- as.matrix(df_model)  # Includes ML5, ML3, ML2, thalach, thal

# Define the binary outcome variable (convert to numeric for glmnet)
y <- as.numeric(as.character(heartDis_cleaned$target))  # Ensure 0/1 coding

# Verify matrix structure
dim(X)
table(y)  # Check class balance

# Fit LASSO logistic regression with cross-validation
lasso_fit_bin <- cv.glmnet(X, y, alpha = 1, family = "binomial")

# Plot the cross-validation error
plot(lasso_fit_bin)

# Extract the best lambda (penalty parameter)
best_lambda_bin <- lasso_fit_bin$lambda.1se
cat("Optimal lambda (binary outcome):", best_lambda_bin, "\n")

# Get LASSO coefficients at the optimal lambda
lasso_coefs_bin <- coef(lasso_fit_bin, s = "lambda.min")
print(lasso_coefs_bin)
```

### Predict Probabilities:

```{r}
# Predict probabilities
predicted_probs <- predict(lasso_fit_bin, newx = X, s = "lambda.1se", type = "response")

# Convert to binary classifications (threshold = 0.5)
predicted_classes <- ifelse(predicted_probs > 0.5, 1, 0)

# Load required package
library(caret)

# Convert `y` (actual values) to factor to match `predicted_classes`
y_factor <- factor(y, levels = c(0,1))
predicted_factor <- factor(predicted_classes, levels = c(0,1))

# Generate the confusion matrix
conf_matrix <- confusionMatrix(predicted_factor, y_factor, positive = "1")

# Print the confusion matrix
print(conf_matrix)
```
